\documentclass{article}
\usepackage[osf,sc]{mathpazo}
\usepackage[utf8]{inputenc}
\usepackage{microtype}

\usepackage{multicol}
\usepackage{pdflscape}
\usepackage[left=1in, right=1in, top=1in, bottom=1in]{geometry}

\usepackage{enumitem}
\usepackage{setspace}

\setlength\columnsep{10pt}
\setlength\itemsep{0pt}
\setlist[itemize]{leftmargin=1.5pc}
\renewcommand\labelitemi{$\circ$}
\setcounter{secnumdepth}{0}

\newcommand{\godcategory}[1]{\bigskip\noindent \textbf{#1}\smallskip\hrule\nopagebreak}
\newcommand{\Lord}{\textsc{Lord}}
\newcommand{\elips}{\thinspace\ldots}
\newcommand{\fivesixtytwo}[1]{\textsc{five hundred sixty-two press, {\romannumeral #1}}}

\begin{document}
\vspace*{16em}

\hrule\smallskip
\begin{center}
{\huge
  \hfill Close Reading \emph{\&} Comparative Assignment \hfill
}

\vspace*{1.2em}

{\LARGE
  Soren Bjornstad
}

\vspace{2.2em}

{\Huge
  $\sim$
}

\vspace{1.5em}

{\Large
  \textsc{rel} 222: The Biblical God

  \vspace{0.3em}

  Prof.\ David Kaden

  \vspace{0.5em}

  24 March 2015
}
\end{center}
\smallskip\hrule

\vspace{3em}

\tableofcontents

\vfill
\begin{center}\fivesixtytwo{2015}\end{center}

\newgeometry{left=0.75in, right=0.75in, top=0.65in, bottom=0.65in}
\begin{landscape}
\begin{multicols}{3}
\section{Actions of God: Judges}
\godcategory{God speaks.}
\begin{itemize}
  \item The \Lord\ \emph{said}, “Judah shall go up. I hereby give the land into his hand.”~(1:2)
  \item Now the angel of the \Lord\ went up from Gilgal to Bochim, and \emph{said}, “I brought you up from Egypt, and brought you into the land that I had promised to your ancestors. \emph{I said}, “I will never break my covenant with you.”~(2:1)
  \item \emph{So now I say}, I will not drive them out before you; but they shall become adversaries to you, and their gods shall be a snare to you.”~(2:3)
  \item When the angel of the \Lord\ \emph{spoke these words} to all the Israelites~(2:4)
  \item So the anger of the \Lord\ was kindled against Israel; \emph{and he said}, “Because this people have transgressed my covenant that I commanded their ancestors, and have not obeyed my voice\elips”~(2:20)
  \item The \Lord\ sent a prophet to the Israelites, and he \emph{said} to them, “Thus says the \Lord, the God of Israel: I led you up from Egypt, and brought you out of the house of slavery\elips~(6:8)
  \item And \emph{I said to you}\elips~(6:10)
  \item The angel of the \Lord\ appeared to him \emph{and said to him}, “The \Lord\ is with you, you mighty warrior.”~(6:12)
  \item The \Lord\ \emph{said to him}, “But I will be with you\elips''~(6:16)
  \item \emph{And he said}, “I will stay until you return.”~(6:18)
  \item \emph{The angel of God said} to him\elips~(6:20)
  \item But the \Lord\ \emph{said to him}, “Peace be to you; do not fear, you shall not die.”~(6:23)
  \item That night the \Lord\ \emph{said to him}\elips~(6:25)
  \item Then I shall know that you will deliver Israel by my hand, \emph{as you have said}.~(6:37)
  \item The \Lord\ \emph{said to Gideon}\elips~(7:2)
  \item Then the \Lord\ \emph{said to Gideon}, “The troops are still too many; take them down to the water, and I will sift them out for you there. When \emph{I say}\elips when \emph{I say}\elips''~(7:4)
  \item And the \Lord\ \emph{said to Gideon}\elips~(7:5)
  \item Then the \Lord\ \emph{said to Gideon}, “With the three hundred that lapped I will deliver you, and give the Midianites into your hand.''~(7:7)
  \item That same night the \Lord\ \emph{said to him}, “Get up, attack the camp, for I have given it into your hand.”~(7:9)
  \item And the angel of the \Lord\ appeared to the woman \emph{and said to her}, “Although you are barren, having borne no children, you shall conceive and bear a son.”~(13:3)
  \item But he \emph{said to me}, “You shall conceive and bear a son.”~(13:7)
  \item “Are you \emph{the man who spoke} to this woman?” (13:11)
  \item The angel of the \Lord\ \emph{said}\elips~(13:13)
  \item The angel of the \Lord\ \emph{said to Manoah}\elips~(13:16)
  \item But the angel of the \Lord\ \emph{said to him}\elips~(13:18)
\end{itemize}

\godcategory{God listens and answers.}
\begin{itemize}
  \item \ldots so that God \emph{may listen} to you.~(9:7)
  \item And the \Lord\ answered\elips~(20:18)
  \item The \Lord\ answered\elips~(20:28)
\end{itemize}

\godcategory{God commands.}
\begin{itemize}
  \item \ldots the commandments of the \Lord, which he \emph{commanded their ancestors} by Moses.~(3:4)
  \item The \Lord, the God of Israel, commands you.~(4:6)
\end{itemize}

\godcategory{God intervenes in war and gives peoples into each others' hands.}
\begin{itemize}
  \item The \Lord\ said, “Judah shall go up. \emph{I hereby give the land into his hand}.”~(1:2)
  \item Then Judah went up and the \Lord\ \emph{gave the Canaanites and the Perizzites into their hand}.~(1:4)
  \item So the anger of the \Lord\ was kindled against Israel, and he \emph{gave them over to plunderers} who plundered them, and he \emph{sold them into the power of their enemies} all around, so that they could no longer withstand their enemies.~(2:14)
  \item Whenever they marched out, the hand of the \Lord\ \emph{was against them to bring misfortune}, as the \Lord\ had warned them and sworn to them, and they were in great distress.~(2:15)
  \item The spirit of the \Lord\ came upon him, and he judged Israel; he went out to war, and the \Lord\ \emph{gave King Cushan-rishathaim of Aram into his hand}.~(3:1)
  \item For the \Lord\ \emph{has given your enemies the Moabites into your hand}.~(3:28)
  \item So the \Lord\ \emph{sold them into the hand of King Jabin of Canaan}.~(4:2)
  \item \emph{I will draw out Sisera}, the general of Jabin’s army\elips \emph{I will give him into your hand}.~(4:7)
  \item For the \Lord\ \emph{will sell Sisera} into the hand of a woman.~(4:9)
  \item Then Deborah said to Barak, “Up! For this is the day on which the \Lord\ \emph{has given Sisera into your hand}.''~(4:14)
  \item And the \Lord\ \emph{threw Sisera and all his chariots and all his army into a panic} before Barak.~(4:15)
  \item So on that day God \emph{subdued King Jabin of Canaan} before the Israelites.~(4:23)
  \item And the \Lord\ \emph{gave them into the hand of Midian} seven years.~(6:1)
  \item And \emph{I delivered you from the hand of the Egyptians}, and from the hand of all who oppressed you, and \emph{drove them out before you}, and \emph{gave you their land.}~(6:9)
  \item “Get up, for the \Lord\ \emph{has given the army of Midian into your hand}.”~(7:15)
  \item That same night the \Lord\ said to him, “Get up, attack the camp, for \emph{I have given it into your hand.}”~(7:9)
  \item The \Lord\ \emph{set every man’s sword against his fellow} and against all the army.~(7:22)
  \item God \emph{has given into your hands} the captains of Midian, Oreb and Zeeb.~(8:3)
  \item Gideon replied, “Well then, when the \Lord\ \emph{has given Zebah and Zalmunna into my hand}\elips~(8:7)
  \item If\elips the \Lord\ \emph{gives them over to me}, I will be your head.~(11:9)
  \item Then the \Lord, the God of Israel, \emph{gave Sihon and all his people into the hand of Israel.}~(11:21)
  \item So now the \Lord, the God of Israel, \emph{has conquered the Amorites} for the benefit of his people Israel.~(11:23)
  \item And should we not be the ones to possess everything that the \Lord\ our God \emph{has conquered} for our benefit?~(11:24)
  \item And the \Lord\ gave them into his hand.~(11:32)
  \item And the \Lord\ gave them into my hand.~(12:3)
  \item And the \Lord\ gave them into the hand of the Philistines forty years.~(13:1)
  \item You \emph{have granted this great victory} by the hand of your servant.~(15:18)
  \item God has indeed given it [the land] into your hands.~(18:10)
  \item The \Lord\ \emph{defeated Benjamin} before Israel.~(20:35)
\end{itemize}

\godcategory{God delivers the Israelites, or chooses not to.}
\begin{itemize}
  \item So now I say, “\emph{I will not drive them out before you}; but they shall become adversaries to you, and their gods shall be a snare to you.”~(2:3)
  \item Whenever the \Lord\ raised up judges for them, the \Lord\ was with the judge, and he \emph{delivered them from the hand of their enemies} all the days of the judge.~(2:18)
  \item “\emph{I will no longer drive out before them} any of the nations that Joshua left when he died.”~(2:21)
  \item And \emph{I delivered you from the hand of the Egyptians}, and from the hand of all who oppressed you, and drove them out before you, and gave you their land.~(6:9)
  \item Then I shall know that \emph{you will deliver Israel by my hand}, as you have said.~(6:37)
  \item Then the \Lord\ said to Gideon, “With the three hundred that lapped \emph{I will deliver you}, and give the Midianites into your hand. Let all the others go to their homes.”~(7:7)
  \item Yet you have abandoned me and worshiped other gods; \emph{therefore I will deliver you no more}.~(10:13)
\end{itemize}

\godcategory{God brings the Israelites into and gives them land.}
\begin{itemize}
  \item The \Lord\ was with Judah, and he \emph{took possession of} the hill country.~(1:19)
  \item \emph{Hebron was given to Caleb}, as Moses had said.~(1:20)
  \item And they abandoned the \Lord, the God of their ancestors, \emph{who had brought them out of the land of Egypt}.~(2:12)
  \item Thus says the \Lord, the God of Israel: \emph{I led you up from Egypt}, and brought you out of the house of slavery.~(6:8)
  \item And I\elips \emph{gave you their land}.~(6:9)
\end{itemize}

\godcategory{The spirit of God comes upon people.}
\begin{itemize}
  \item The spirit of the \Lord\ \emph{came upon him}, and he judged Israel; he went out to war, and the \Lord\ gave King Cushan-rishathaim of Aram into his hand.~(3:1)
  \item But the spirit of the \Lord\ \emph{took possession of} Gideon.~(6:34)
  \item Then the spirit of the \Lord\ \emph{came upon} Jephthah.~(11:29)
  \item The spirit of the \Lord\ \emph{began to stir him} in \hbox{Mahaneh-dan.}~(13:25)
  \item The spirit of the \Lord\ rushed on him.~(14:6)
  \item Then the spirit of the \Lord\ rushed on him.~(14:19)
  \item The spirit of the \Lord\ rushed on him.~(15:14)
\end{itemize}

\clearpage

\godcategory{God strengthens people.}
\begin{itemize}
  \item And the \Lord\ \emph{strengthened King Eglon} of Moab against Israel, because they had done what was evil in the sight of the \Lord.~(3:12)
  \item Then Samson called to the \Lord\ and said, “\Lord\ God, remember me and strengthen me only this once, O God, so that with this one act of revenge I may pay back the Philistines for my two eyes.” [It is implied that God then does this.]~(16:28)
\end{itemize}

\godcategory{God is with people (or not).}
\begin{itemize}
  \item The house of Joseph also went up against Bethel; and the \Lord\ \emph{was with them}.~(1:22)
  \item The \Lord\ \emph{had left those nations}, not driving them out at once, and had not handed them over to Joshua.~(2:23)
  \item And he said, “\emph{I will stay until you return}.”~(6:18)
  \item But he did not know that the \Lord\ \emph{had left him}.~(16:20)
\end{itemize}

\godcategory{God blesses and curses.}
\begin{itemize}
  \item \emph{Curse Meroz}, says the angel of the \Lord, \emph{curse bitterly its inhabitants}.~(5:23)
  \item The \Lord\ blessed him [Samson].~(13:24)
  \item May my son be blessed by the \Lord!~(17:2)
  \item Now I know that the \Lord\ \emph{will prosper me}, because the Levite has become my priest.~(17:13)
\end{itemize}

\godcategory{God travels, appears to people, and vanishes again.}
\begin{itemize}
  \item The \Lord\ is indeed \emph{going out before you}.~(4:14)
  \item \Lord, \emph{when you went out from Seir}, \emph{when you marched} from the region of Edom\elips~(5:4)
  \item Now the angel of the \Lord\ \emph{came and sat under the oak} at Ophrah.~(6:11)
  \item The angel of the \Lord\ \emph{appeared to him} and said to him, “The \Lord\ is with you, you mighty warrior.”~(6:12)
  \item Then the \Lord\ \emph{turned to him}.~(6:14)
  \item And the angel of the \Lord\ \emph{vanished from his sight}.~(6:21)
  \item Then the woman came and told her husband, “\emph{A man of God came to me}.”~(13:6)
  \item God listened to Manoah, and the angel of God \emph{came again} to the woman as she sat in the field.~(13:9)
  \item The [angel of God] who came to me the other day \emph{has appeared to me}.~(13:10)
  \item The angel of the \Lord\ \emph{did not appear again} to Manoah and his wife.~(13:21)
\end{itemize}

\godcategory{God chooses judges and rulers for the people.}
\begin{itemize}
  \item Then the \Lord\ raised up judges.~(2:16)
  \item Whenever the \Lord\ \emph{raised up judges for them}, the \Lord\ was with the judge.~(2:18)
  \item The \Lord\ raised up a deliverer for the Israelites.~(3:9)
  \item The \Lord\ raised up for them a deliverer.~(3:15)
  \item The \Lord\ sent a prophet to the Israelites.~(6:8)
\end{itemize}

\godcategory{God tests.}
\begin{itemize}
  \item In order to test Israel\elips~(2:22)
  \item Now these are the nations that the \Lord\ left \emph{to test all those in Israel} who had no experience of any war in Canaan.~(3:1)
  \item They were \emph{for the testing of Israel}, to know whether Israel would obey the commandments of the \Lord.~(3:4)
\end{itemize}

\godcategory{God gives signs.}
\begin{itemize}
  \item Then the angel of the \Lord\ reached out the tip of the staff that was in his hand, and touched the meat and the unleavened cakes; and fire sprang up from the rock and consumed the meat and the unleavened cakes; and the angel of the \Lord\ vanished from his sight.~(6:21)
  \item I am going to lay a fleece of wool on the threshing floor; if there is dew on the fleece alone, and it is dry on all the ground, then I shall know that you will deliver Israel by my hand, as you have said.~(6:37)
  \item So God split open the hollow place that is at Lehi, and water came from it.~(15:19)
\end{itemize}

\godcategory{God rules and judges.}
\begin{itemize}
  \item Then the \Lord\ said to Gideon, “The troops are still too many; take them down to the water, and I will sift them out for you there.~(7:4)
  \item The \Lord\ \emph{will rule over you.}~(8:23)
  \item And the elders of Gilead said to Jephthah, “The \Lord\ \emph{will be witness between us}.”~(11:10)
  \item Let the \Lord, \emph{who is judge, decide today} for the Israelites or for the Ammonites.~(11:27)
\end{itemize}

\clearpage

\godcategory{God gives offspring.}
\begin{itemize}
  \item And the angel of the \Lord\ appeared to the woman and said to her, “Although you are barren, having borne no children, \emph{you shall conceive and bear a son}.”~(13:3)
  \item For you shall conceive and bear a son.~(13:5)
  \item But he said to me, “You shall conceive and bear a son.”~(13:7)
\end{itemize}

\godcategory{God accepts sacrifices.}
\begin{itemize}
  \item The angel of the \Lord\ ascended in the flame of the altar.~(13:20)
  \item If the \Lord\ had meant to kill us, he would not have \emph{accepted a burnt offering and grain offering} at our hands, or shown us all these things, or now announced to us such things as these.~(13:23)
\end{itemize}

\godcategory{God feels anger and pity.}
\begin{itemize}
  \item So the anger of the \Lord\ was kindled against Israel.~(2:14)
  \item For the \Lord\ \emph{would be moved to pity} by their groaning.~(2:18)
  \item So the anger of the \Lord\ was kindled against Israel.~(2:20)
  \item Therefore the anger of the \Lord\ was kindled against Israel.~(3:8)
\end{itemize}

\godcategory{Miscellaneous}

\godcategory{God pays people back.}
\begin{itemize}
  \item As I have done, so God has paid me back.~(1:7)
\end{itemize}

\godcategory{God makes covenants.}
\begin{itemize}
  \item {[The angel of the \Lord]} said, ``I will never break my covenant with you.’’~(2:1)
\end{itemize}

\godcategory{God causes fear.}
\begin{itemize}
  \item The mountains quaked before the \Lord.~(5:5)
\end{itemize}

\godcategory{God sends evil spirits.}
\begin{itemize}
  \item But God \emph{sent an evil spirit} between Abimelech and the lords of Shechem.~(9:23)
\end{itemize}

\godcategory{God watches over people.}
\begin{itemize}
  \item The mission you are on \emph{is under the eye} of the \Lord.~(18:6)
\end{itemize}

\godcategory{God creates breaches.}
\begin{itemize}
  \item \ldots because the \Lord\ had \emph{made a breach} in the tribes of Israel.~(21:15)
\end{itemize}


\clearpage

\section{Actions of God: 2 Thessalonians}
\godcategory{God gives grace, peace, and relief.}
\begin{itemize}
  \item \emph{Grace to you and peace} from God our Father and the Lord Jesus Christ.~(1:2)
  \item \ldots and \emph{to give relief to the afflicted} as well as to us, when the Lord Jesus is revealed from heaven with his mighty angels~(1:7)
  \item Now may the Lord of peace himself \emph{give you peace} at all times in all ways.~(3:16)
  \item The \emph{grace of our Lord Jesus Christ be with all of you.}~(3:18)
\end{itemize}

\godcategory{God judges and does justice and destroys evil.}
\begin{itemize}
  \item This is evidence of the \emph{righteous judgment of God}.~(1:5)
  \item For it is indeed just of God to \emph{repay with affliction} those who afflict you.~(1:6)
  \item \ldots in flaming fire, \emph{inflicting vengeance} on those who do not know God and on those who do not obey the gospel of our Lord Jesus.~(1:8)
  \item These \emph{will suffer the punishment of eternal destruction}, separated from the presence of the Lord and from the glory of his might.~(1:9)
  \item And then the lawless one will be revealed, whom the Lord Jesus \emph{will destroy} with the breath of his mouth, \emph{annihilating him} by the manifestation of his coming.~(2:8)
\end{itemize}

\godcategory{God chooses and calls you.}
\begin{itemize}
  \item To this end we always pray for you, asking that our God \emph{will make you worthy of his call}.
  \item \ldots because God \emph{chose you} as the first fruits.~(2:13)
  \item For this purpose he \emph{called you}.~(2:14)
  \item May the Lord \emph{direct your hearts} to the love of God and to the steadfastness of Christ.~(3:5)
\end{itemize}

\godcategory{God comes and is revealed.}
\begin{itemize}
  \item \ldots and to give relief to the afflicted as well as to us, \emph{when the Lord Jesus is revealed} from heaven with his mighty angels.~(1:7)
  \item \ldots \emph{when he comes to be glorified} by his saints and to be marveled at on that day among all who have believed\elips~(1:10)
  \item As to \emph{the coming of} our Lord Jesus Christ\elips~(2:1)
  \item And then \emph{the lawless one will be revealed}, whom the Lord Jesus will destroy with the breath of his mouth, annihilating him \emph{by the manifestation of his coming}.~(2:8)
\end{itemize}

\godcategory{God deludes the unfaithful.}
\begin{itemize}
  \item For this reason God \emph{sends them a powerful delusion}, leading them to believe what is false.~(2:11)
\end{itemize}

\godcategory{God fulfills things which ought to be.}
\begin{itemize}
  \item To this end we always pray for you, asking that our God \emph{will\elips fulfill by his power} every good resolve and work of faith.~(1:11)
\end{itemize}

\godcategory{God reveals.}
\begin{itemize}
  \item That day will not come unless the rebellion comes first and \emph{the lawless one is revealed}, the one destined for destruction.~(2:3)
  \item And then \emph{the lawless one will be revealed}, whom the Lord Jesus will destroy with the breath of his mouth, annihilating him by the manifestation of his coming.~(2:8)
\end{itemize}

\godcategory{God strengthens the believer to do good things.}
\begin{itemize}
  \item Now may our Lord Jesus Christ himself and God our Father, who loved us and through grace gave us eternal comfort and good hope, comfort your hearts and \emph{strengthen them in every good work} and word.~(2:16)
  \item But the Lord is faithful: he \emph{will strengthen you and guard you from the evil one}.~(3:3)
\end{itemize}

\end{multicols}
\end{landscape}
\restoregeometry



\section{Conclusions}
\large
\setstretch{1.9}

Throughout many parts of the Bible, God as a character guides events on earth. The ways in which he chooses to do this are many and varied, and the methods change significantly between the Old and New Testaments, including between Judges and 2 Thessalonians.

In the Old Testament book of Judges, God intervenes directly in his people's affairs in a great variety of ways, taking actions that have outwardly visible effects in the present. God communicates with his people, telling them what to do, listening to their concerns, and in some cases coming to visit them. Further, he influences the outcome of many battles and other events. In 2 Thessalonians, in contrast, God no longer takes an active, visible role in the development of the story. However, God is still there; in particular, his spirit is present with believers and provides them emotional support, and he calls them to be loving and do good things. God also continues to judge, and may take action to right wrongs, such as \emph{repaying with affliction}. But more importantly, 2 Thessalonians talks incessantly about the future, a future in which God will again be more visible: God will come again, be revealed, and destroy evil. And even now, all people will have to face God's judgment and potential punishment when they die.

In Judges, God's character appeals to a sense of justice and righteousness, with respect to the ``right'' people, the Israelites. While indeed he only favors those people, and often deals quite harshly with others, when he favors them, he does so because they have acted rightly, and when he is angry, it is because they have (like clockwork) ``again done what was evil in the sight of the Lord.'' When God sends evil spirits or causes destruction, he does it against the ``wrong'' people in order to accomplish something for the Israelites. In 2 Thessalonians, the dichotomy between the ``right'' and ``wrong'' people changes to \emph{believers vs.\ unbelievers}, which is somewhat more comfortable for the modern reader: the people who gain God's favor and avoid his wrath were not merely in the right place at the right time, they did something to deserve it (they chose to believe).

The impressions of God presented in these two books are not inconsistent with each other. However, they certainly present a God who has changed, who chooses to act in different ways in more recent times.

\end{document}
